### 2021-10-12 chapters/introduction.tex
Removed unused comments

\chapter*{Abstract}
\endgroup
\addcontentsline{toc}{chapter}{Abstract} 
% Paper 1
% Here we introduce a hierarchical model trained on numerical-relativity simulations, which can generate reliable post-merger spectra in a fraction of a second. Our spectra have mean fitting factors of 0.95, which compares to a fitting factor of 0.93 between different numerical-relativity codes that simulate the same physical system. This method is the first step towards generating large template banks of spectra for use in post-merger detection and parameter estimation.
% Paper 2
% Here we develop a fast, simple model that can generate gravitational waveforms, and show it can be used for both detection and parameter estimation of post-merger remnants.
% The model consists of three exponentially-damped sinusoids with a linear frequency-drift term.
% We test the model against nine equal-mass numerical-relativity simulations selected for emission of gravitational waves for $\gtrsim25$\,ms.
% The median fitting factors between the model waveforms and numerical-relativity simulations exceed 0.90.
% We detect remnants at a post-merger signal-to-noise ratio of $\ge 7$ using a Bayes-factor detection statistic with a threshold of 3000.
% We can constrain the primary post-merger frequency to $\pm_{1.2}^{1.4}\%$ at post-merger signal-to-noise ratios of 15 with an increase in precision to $\pm_{0.2}^{0.3}\%$ for post-merger signal-to-noise ratios of 50.
% The tidal coupling constant can be constrained to $\pm^{9}_{12}\%$ at post-merger signal-to-noise ratios of 15, and $\pm 5\%$ at post-merger signal-to-noise ratios of 50 using a hierarchical inference model.
% Paper 3
% We develop and test a method to measure the collapse time of post-merger remnants.
% We show that for a GW170817-like event at $\sim\!40$\,Mpc, a network of Einstein Telescope with Cosmic Explorer is required to detect collapse times of $\sim\!10$\,ms.
% For a two-detector network at A+ design sensitivity, post-merger remnants with collapse times of $\sim\!10\,\mathrm{ms}$ must be $\lesssim 10$\,Mpc to be measureable.
% This increases to $\sim\!18\Hyphdash*26$\,Mpc if we include the proposed Neutron star Extreme Matter Observatory (NEMO), increasing the effective volume by a factor of $\sim\!30$.
    % Surviving neutron star remnants from binary neutron star mergers are some of the most extreme forms of matter known in the Universe.
    Surviving remnants from binary neutron star mergers, remnants that have not yet collapsed to a black hole,  are some of the most extreme forms of matter known in the Universe.    
    The densities of surviving merger remnants are $\sim 10^{14}-10^{15}\,\mathrm{g\, cm^{-3}}$ which is up to several times nuclear density.
    The temperature of the post-merger remnant can reach values from tens to hundreds of $\mathrm{MeV}$.
    This combination of extreme temperatures and densities result in a state of matter that is observable only in proto-neutron stars and post-merger remnants.
    Although binary neutron star post-merger remnants have yet to be directly observed, they are a target for gravitational-wave detection in the near future. \par

@@

    Gravitational waves are a phenomenon predicted by general relativity which was formulated more than 100 years ago by Albert Einstein with the famous formula
    \begin{align}
        G_{\mu\nu} & = R_{\mu\nu} - \frac{1}{2}R  g_{\mu\nu}  =  \frac{8\pi G}{c^4} T_{\mu\nu}, \label{eq:Intro:GeqT}
    \end{align}
    where $G_{\mu\nu}$ is the space-time curvature tensor and $T_{\mu\nu}$ is the stress energy tensor. The local metric tensor is $g_{\mu\nu}$, $R_{\mu\nu}$ is the Ricci tensor and $R$ is the Ricci scalar. 
    The gravitational constant is $G$ and the speed of light in vacuum is $c$, though for remainder of this work we will use geometric units of $G=c=1$ unless otherwise stated. 
    Equation~\ref{eq:Intro:GeqT} demonstrates how the curvature of the space-time, $G_{\mu\nu}$, alters the dynamics of matter, $T_{\mu\nu}$, and vice-versa. \par
    Movement of matter in space can generate gravitational waves.
    For simulation purposes, the movement of matter can be modelled by the Einstein field equations~Eq.(\ref{eq:Intro:GeqT}) and perturbations of black holes~\cite[e.g.,][]{Moncrief1974,Abrahams1996,Andrade1999,Nagar2005}.
    This methodology allows the calculation of multipole functions of odd ($\mathcal{Q}^\times_{lm}$) and even ($\mathcal{Q}^+_{lm}$) parity, with angular indices of $(l,m)$.
    These multipoles can be evaluated by numerical-relativity simulations which allow the extraction of gravitational waves.
% \redtext{Movement of matter in space generates mass multipoles that can be related to perturbations of black holes.
%     These perturbations can be described in terms of the non-rotating metric for black holes, the Schwarzschild metric, and the Einstein field equations,~Eq.~\ref{eq:Intro:GeqT}.
%     Non-spherical gauge-invariant perturbations, outlined in Ref.~\cite{Baiotti2008}, can generate master multipole functions of odd ($\mathcal{Q}^\times_{lm}$) and even ($\mathcal{Q}^+_{lm}$) parity, with angular indices of $(l,m)$, that can be estimated my numerical-relativity simulations (see also Refs.~\cite{Moncrief1974,Abrahams1996,Andrade1999,Nagar2005}).}
% \redtext{
%     Gravitational waves can be extracted for a virtual observer on the 2-sphere of a Schwarzschild metric with an isotropic radius of $r$ from the source.
%     The Schwarzschild radius for this metric is $r_S=r(1-M/2r)^2$, where $M$ is the enclosed mass~\cite[e.g.,][]{Baiotti2008}.
%     Gravitational waves are extracted through non-spherical gauge-invariant perturbations of the Schwarzschild metric, resulting in multipoles  where $(l,m)$ are the angular indices~\cite{Moncrief1974,Abrahams1996,Andrade1999,Nagar2005,Baiotti2008}.\footnote{Please note that $\mathcal{Q}^+_{lm}$ and $\mathcal{Q}^\times_{lm}$ have different units, see Ref.~\cite{Nagar2005} for more information.}\par  % Matthews1962
%     % Check out: Abrahams, Rezzolla, Rupright, Anderson, Anninos, Baumgarte, et al., Phys. Rev. Lett. 80, 1812 (1998) 
%     % Rupright, Abrahams, Rezzolla, Phys. Rev. D 58, 044005 (1998).
%     % Rezzolla, Abrahams, Matzner, Rupright, Shapiro, Phys. Rev. D 59, 064001 (1999).
%     }
    The plus, $h_{+}$, and cross, $h_{\times}$, polarisations of the gravitational waves are related to the multipoles as follows~\cite{Moncrief1974,Abrahams1996,Andrade1999,Nagar2005,Baiotti2008}:\footnote{Please note that $\mathcal{Q}^+_{lm}$ and $\mathcal{Q}^\times_{lm}$ have different units, see Ref.~\cite{Nagar2005} for more information.}
     \begin{align}
        h_{+} - i h_{\times} & = \frac{1}{r\sqrt{2}}\sum_{l,m} {}_{-2}Y^{lm}\left(\mathcal{Q}^{+}_{lm} -i \int_{-\infty}^t\mathcal{Q}^\times_{lm}(t')dt'\right). \label{eq:Intro:MultipoleExpansion}
    \end{align}   
    Here, $r$ is the distance to the gravitational-wave source and ${}_{-2}Y^{lm}$ are the spin-weighted spherical harmonics with spin $s=-2$ and angular indices of $(l,m)$.
    For most sources, gravitational waves are predominantly emitted in the $l=m=2$ mode. 
    In this case Eq.(\ref{eq:Intro:MultipoleExpansion}) evaluates to
     \begin{align}
        h_{+} + i h_{\times} & = \frac{1}{2 r} \sqrt{\frac{5}{2\pi }} e^{2 i \phi } \cos ^4\left(\frac{\theta }{2}\right)\left(\mathcal{Q}^{+}_{22} -i \int_{-\infty}^t\mathcal{Q}^\times_{22}(t')dt'\right), \label{eq:Intro:MultipoleExpansionlm22}
    \end{align}
    where  $\theta$ is the polar angle and $\phi$ is the azimuthal angle usually defined relative to orbital angular momentum vector. \par

@@

    Assuming a circular orbit of two point masses of mass $M_1$ and $M_2$, the Newtonian quadrupole formula can be used to calculate the evolution of the orbital separation, $a$, due to gravitational wave emission:~\cite[e.g.,][]{Cutler1994}
    \begin{align}
        \frac{da}{dt} & = -\frac{64}{5}\frac{\mu M^2}{a^3}, \label{eq:Intro:OrbitalEvolution}        
    \end{align}
    where  $M \equiv M_1 + M_2$ is the total mass, and $\mu \equiv M_1 M_2 / M$ is the reduced mass.
    The time evolution of gravitational-wave strain frequency can then be calculated as
    \begin{align}
        \frac{df}{dt} & = \frac{96}{5}\pi^{8/3}\mathcal{M}^{5/3}f^{11/3}, \label{eq:Intro:FreqEvolution}        
    \end{align}   
    where $\mathcal{M} \equiv \mu^{3/5} M^{2/5}$ is the chirp mass of the binary system. 
    Using Eqs.(\ref{eq:Intro:OrbitalEvolution}\Hyphdash*\ref{eq:Intro:FreqEvolution}), an expression for the frequency and the gravitational-wave strain magnitude can be calculated~~\cite[e.g.,][]{Cutler1994}: 
      \begin{align}
      f &  \propto \mathcal{M}^{- 5/8}(t_c-t)^{- 3/8},\label{eq:Intro:chirp_f}\\
        % |h(t)| & \propto  \frac{\pi^{2/3}}{D}\left(\frac{576\mu^3 M^2}{5}\right)^{1/4}(t_c-t)^{-1/4}, \label{eq:Intro:chirp_ht_old} ,\\
        |h(t)| & \propto  \frac{\mathcal{M}^{5/4}}{D}(t_c-t)^{-1/4}, \label{eq:Intro:chirp_ht} 
        % Calculated by Cutler1994 (2.12) + (2.11) + Chirp mass defn
    \end{align}
    where $t$ is the time, $t_c$ is the time until coalescence, and $D$ is the luminosity distance to the source. 
    The combination of Eq.(\ref{eq:Intro:chirp_f})~and~Eq.(\ref{eq:Intro:chirp_ht}) result in the chirp waveform detected in the inspiral of compact object mergers.
    \par
    The first direct gravitational-wave detection from the inspiral of a binary neutron star merger, GW170817, was observed by a detector network consisting of the LIGO detectors at Hanford, Washington and Livingston, Louisiana in the USA, and the Virgo detector in Cascina, Italy~\cite{AdvancedLIGO2015,AdvancedVirgo2015,GW170817Detection}. 
    All three detectors consist of `L' shaped laser interferometers with free-falling test-masses.
    The gravitational-wave strain is found by measuring the optical path difference between the perpendicular beam tubes. \par

    Another binary neutron star (possibly neutron star - black hole) merger, GW190425, was detected in 2019~\cite{GW190425Detection,GWTC2,PopGWTC2}.
    %, along with another binary neutron star black hole binary with a high false alarm rate, GW190426\_152155.
    However, the luminosity distance of GW190425, $159^{+69}_{-72}\,$Mpc, and the broad sky localisation of the source, $8284\,\mathrm{deg}^2$, precluded any coincident electromagnetic observations~\cite{GW190425Detection}. In comparison, the sky localisation of GW170817 was $28\,\mathrm{deg}^2$ at a distance of $\sim 40\,\Mpc$.
    Therefore, to date, GW170817 is the only gravitational-wave signal that was observed with coincident electromagnetic radiation and will be discussed further in Section~\ref{sec:Intro:NeutronStarMergers}.
    %, allowing the host galaxy to be located~\cite{GW170817multi}.

@@

% ***************************************************************************************************************************************************************
    \section{Neutron star mergers} \label{sec:Intro:NeutronStarMergers}
    Binary neutron star mergers emit vast amounts of energy in gravitational waves.
    The estimated radiated energy for the GW170817 merger was $\gtrsim 5\times 10^{52}\,$ergs~\cite{GW170817Detection}, equivalent to $\sim 1\%$ of the total rest mass of the two neutron stars.
    For a source at a distance of $D$ from the observer, the amplitude of gravitational waves are  $\propto 1/D$, whereas the amplitude of electromagnetic waves are $\propto 1/D^2$.
    Additionally, electromagnetic waves can be attenuated by intervening media between the source and the detector whereas gravitational waves are transparent to the intervening media.
    Hence gravitational waves are detectable over much larger distances in comparison to electromagnetic radiation with the same source luminosity. \par
    % A sentence here about the lack of opacity to GW as opposed to EM?

    % If a BNS system of Rinit is less than tmerge is left undisturbed then the BNS can merge.
    % Formation channels.
    % If a binary neutron star system is formed with a separation less than the cr
    % From GW170817multi:
    % This motivated the development of accurate models for the two-body, general-relativistic dynamics (Blanchet et al. 1995;Buonanno & Damour 1999; Pretorius 2005; Baker et al. 2006; Campanelli et al. 2006; Blanchet 2014) that

    % Damour2012 in terms of tidal coupling constant, $\kappa_2^T$
    Gravitational-wave strain emitted from the inspiral of binary neutron star mergers, approximated by Eqs.(\ref{eq:Intro:FreqEvolution}\Hyphdash*\ref{eq:Intro:chirp_ht}), can be recast in the frequency domain using the stationary phase approximation~\cite{Cutler1994}
    \begin{align}
        h_{22}(f) & \equiv \frac{\mathcal{A}}{D}\mathcal{M}^{5/6}f^{-7/6} \exp(i\Psi(f)), \label{eq:Intro:StationaryPhaseApproxH22} \\
        \Psi(f) & = 2\pi f t_c - \phi_c - \frac{\pi}{4} +\frac{3}{4}\left(8\pi\mathcal{M}f\right)^{-5/3}, \label{eq:Intro:StationaryPhaseApproxPsi} 
    \end{align}
    where $h_{22}(f)$ is the Fourier transform of the $(l,m)=(2,2)$ gravitational-wave strain, and $\Psi(f)$ is the corresponding phase.
    Here, $f$ is the frequency~(Eq.(\ref{eq:Intro:FreqEvolution})), $\phi_c$ and $t_c$ are the phase and time of coalescence, and $\mathcal{A}$ is a factor that depends on the orientation of the system.\footnote{The orientation factor, $\mathcal{A}$, depends on angles intrinsic and extrinsic to the binary system. Intrinsic angles include the angles defining the relative orientation of the neutron star spins to the orbital plane. Extrinsic angles include the angles defining the relative orientation between the orbital plane and each of the detectors.}
    Equations~\ref{eq:Intro:StationaryPhaseApproxH22}\Hyphdash*\ref{eq:Intro:StationaryPhaseApproxPsi} are valid for $a \gtrsim 6M$ where $a$ is the orbital separation of the binary system and $a\sim 6M$ is an approximation of the innermost stable circular orbit~\cite{Cutler1994}.
    Solving Eqs.(\ref{eq:Intro:OrbitalEvolution}\Hyphdash*\ref{eq:Intro:FreqEvolution}) for $a \sim 6M$ gives $f_{max}\approx c^3 ( 6^{3/2}\pi G M )^{-1}\,\mathrm{Hz}$.
    \par
    % The base post-Newtonian expansion parameter is $\sqrt{M/r}$ but it is more convenient to use $\left(\pi M f\right)^{1/3}$ as the post-Newtonian expansion parameter.
    % These two parameters are equal up to and excluding terms of order $(M/r)^{3/2}$.
    The validity for these equations can be extended to times closer to the merger by adding relativistic corrections calculated by post-Newtonian expansion and effective-one-body dynamics~\cite{Cutler1994,Blanchet1995, Buonanno1999,Droz1999,Damour2012,Yagi2013,Blanchet2014}. % to Eq.~\ref{eq:Intro:StationaryPhaseApproxPsi}
    These corrections introduce dependence on the masses, spins, and tidal deformabilities of the individual neutron stars.
    This allows the measurement of these system properties in addition to the chirp mass, given a gravitational-wave inspiral detection.
    However, the accuracy of the individual neutron star properties is significantly less than the accuracy of the chirp mass because the relativistic corrections required to measure these properties are fairly weak. 
    This can be seen by looking at the parameters measured from GW170817.

@@

    % Metzger2018: Blue KN eject proposed cause: neutrino-irradiated, magnetically driver outflow from temporarily stable HMNS consistent with $\tc \sim 0.1-1\,\s$
    % Gill2019: Assume jet powered by rotating BH with $M_{\mathrm{ej,blue}} \approx 0.025\,\msun$ then $\tc \sim 1\,\s$
    % Murguia-Berthier2021: Assume jet powered by rotating BH: $\tc \approx1-1.7\,\s$
    % Yu2018: long lived stable NS

    % From Zhang2019:
    % There are two explanations for the total delay time of $\sim 1.7\,$s of GRB 170817A after the merger of GW170817~\cite{Zhang2019}.
    % The first involves a fireball where the emission is most likely from the photosphere radius. 
    % Assuming that the delay time from breakout to GRB emission is $\sim\,2\s$, similar to the duration of the GRB emission, the energy estimate for GRB 170817A suggests a Lorentz factor of $\sim 4$.
    % However, radio afterglow modelling suggests that the Lorentz factor should be $\gg 4$ for $t < 10\,$days after the merger. % Zhang2019[70]
    % This implies that the delay time from breakout to GRB emission is $\ll 2\,\s$ and that the total delay time is mostly attributed to the time to launch a jet or the time for breakout to occur. 
    % This scenario can still work if a post-merger remnant survives $\sim\,1\s$ before launching a jet.
    % However, this means that similarity between the total delay time of $\sim\,1.7\s$ and the GRB burst durations of $\sim\,2\s$ are coincidental as they are produced by different mechanisms. \par
    % The second explanation assumes that the majority of the total delay time is accounted by the time it takes from the breakout of the jet to the emission of the GRB~\cite{Zhang2018}.
    % This can be estimated by the width of the first pulse in GRB 170817A which was $\sim 1\,\s$ but could have been longer due to the weak nature of the burst.
    % This scenario does not depend on the type of post-merger remnant which may be either a neutron star or a black hole.

    % [Go through above references in the first blue paragraph]}

%     \begin{itemize}
%         \item tidal/inspiral methods
%         \item GW170817
%         \item other detections
%         \item Merger outcomes

%     \end{itemize}

% ***************************************************************************************************************************************************************
    \section{Numerical relativity simulations}

    Numerical-relativity simulations are required for determining the dynamics of the late inspiral of the binary neutron star and the post-merger remnant.
    To perform a numerical-relativity simulation, the equation of state must be defined, a set of initial conditions must be evaluated in general relativity, and then Einstein's field equations must be evolved.
    Each of these three requirements make these simulations difficult.
    The equation of state for a `normal' neutron star is still unknown, and the equation of state for a hot post-merger remnant is even more uncertain, after all, this is the hottest and densest bulk matter in the Universe.

    \par
% Good overall references: Daverio2018, Dietrich2016

% ******************************************************************************************************************************************************    
    % Thesis: Maione2017
    % Two different conformal Z4 formulations have been studied: the CCZ4 formulation [Alic2012, Alic2013], which retains the full equations and in this way is a covariant formulation, at the price of greater differences respect to BSSN-OK and the need to introduce an additional parameter k3 to be able to evolve also spacetimes containing black holes and the Z4c [Gundlach2005, Bernuzzi2010, Hilditch2013] formulation, which retains only the principal part of the equations, in order to be as close as possible to BSSN-OK, but is not fully covariant and needs the introduction of ad-hoc constraint preserving boundary conditions in order to avoid constraint violations being reflected back inside the grid at the outer boundaries [Ruiz2011]. 
% ******************************************************************************************************************************************************    
    The initial conditions for binary neutron star mergers in general relativity are complex and a number of methods have been developed for this task.
    Codes that have been used for this include \texttt{LORENE}~\cite{Gourgoulhon2001} and \texttt{COCAL}~\cite{Tsokaros2015}.
    The \texttt{LORENE} formulation assumes a conformally-flat approximation for  neutron star binaries and uses spectral decomposition to solve for the initial conditions~\cite{Gourgoulhon2001}.
    This formulation constructs quasiequilibrium configurations which are approximations of the inspiral dynamics using exactly circular orbits.
    Quasiequilibrium configurations are also generated by \texttt{COCAL}.
    This is performed by excising regions around the neutron stars and solving Green's integral formula using multipole expansion and finite-difference methods~\cite{Uryu2012,Tsokaros2015}.
    % Tsokaros2016: also KADATH [36], and SCRID [37,38]
    Once the initial conditions have been approximated then the space-time can be evolved.

    Space-time evolution methods include the following formulations: Arnowitt-Deser-Misner (ADM)~\cite{Arnowitt2008}\footnote{The original book reference for ADM formalism is out of print and the authors have listed the cited paper as an `intentially unretouched' alternative.},  Baumgarte-Shapiro-Shibata-Nakamura-Oohara-Kojima (BSSNOK)~\cite{Nakamura1987,Shibata1995,Baumgarte1999,Alcubierre2000} and Z4c~\cite{Bona2003,Bona2004,Gundlach2005,Bernuzzi2010,Hilditch2013}. 
    %, and conformally flat (Isenberg-Wilson-Matthews)~\cite{Isenberg2008}. %Conformal and Covariant Z4 (CCZ4)~\cite{Bona2003,Bona2004,Alic2012,Alic2013},  \par
    The ADM formalism is a Hamiltonian method that splits the Einstein's field equations into separate time and space components.
    This is achieved by applying a gauge that consists of a shift vector and lapse function~\cite{Arnowitt2008}.
    Unfortunately, ADM numerical simulations were found to be unstable to accumulated errors over the simulation time~\cite{Brandt2000,Kidder2001,Alcubierre2001a} and a number of different formulations were developed to improve this. \par

@@

    % Baiotti2017: A perhaps disappointing note that needs to be made, however, is that despite this abundance of numerical codes, systematic comparisons of them are seldom done, and the only cases we are aware of are the quantitative comparisons between the SACRA and Whisky codes [160, 490] and between the BAM and SpEC codes [110]; such comparisons are tedious and computationally expensive but important for gauging the systematic errors that the various codes can introduce.

% ***************************************************************************************************************************************************************
    \section{Post-merger remnants} \label{sec:Intro:PMR}
%     \begin{itemize}
%         \item Supramassive, Hypermassive properties
%         \item observational methods, 
%     \end{itemize}

>>>>

\chapter*{Abstract}
\endgroup
\addcontentsline{toc}{chapter}{Abstract}

    Surviving remnants from binary neutron star mergers, remnants that have not yet collapsed to a black hole,  are some of the most extreme forms of matter known in the Universe.    
    The densities of surviving merger remnants are $\sim 10^{14}-10^{15}\,\mathrm{g\, cm^{-3}}$ which is up to several times nuclear density.
    The temperature of the post-merger remnant can reach values from tens to hundreds of $\mathrm{MeV}$.
    This combination of extreme temperatures and densities result in a state of matter that is observable only in proto-neutron stars and post-merger remnants.
    Although binary neutron star post-merger remnants have yet to be directly observed, they are a target for gravitational-wave detection in the near future. \par

@@

    Gravitational waves are a phenomenon predicted by general relativity which was formulated more than 100 years ago by Albert Einstein with the famous formula
    \begin{align}
        G_{\mu\nu} & = R_{\mu\nu} - \frac{1}{2}R  g_{\mu\nu}  =  \frac{8\pi G}{c^4} T_{\mu\nu}, \label{eq:Intro:GeqT}
    \end{align}
    where $G_{\mu\nu}$ is the space-time curvature tensor and $T_{\mu\nu}$ is the stress energy tensor. The local metric tensor is $g_{\mu\nu}$, $R_{\mu\nu}$ is the Ricci tensor and $R$ is the Ricci scalar. 
    The gravitational constant is $G$ and the speed of light in vacuum is $c$, though for remainder of this work we will use geometric units of $G=c=1$ unless otherwise stated. 
    Equation~\ref{eq:Intro:GeqT} demonstrates how the curvature of the space-time, $G_{\mu\nu}$, alters the dynamics of matter, $T_{\mu\nu}$, and vice-versa. \par
    Movement of matter in space can generate gravitational waves.
    For simulation purposes, the movement of matter can be modelled by the Einstein field equations~Eq.(\ref{eq:Intro:GeqT}) and perturbations of black holes~\cite[e.g.,][]{Moncrief1974,Abrahams1996,Andrade1999,Nagar2005}.
    This methodology allows the calculation of multipole functions of odd ($\mathcal{Q}^\times_{lm}$) and even ($\mathcal{Q}^+_{lm}$) parity, with angular indices of $(l,m)$.
    These multipoles can be evaluated by numerical-relativity simulations which allow the extraction of gravitational waves.
    The plus, $h_{+}$, and cross, $h_{\times}$, polarisations of the gravitational waves are related to the multipoles as follows~\cite{Moncrief1974,Abrahams1996,Andrade1999,Nagar2005,Baiotti2008}:\footnote{Please note that $\mathcal{Q}^+_{lm}$ and $\mathcal{Q}^\times_{lm}$ have different units, see Ref.~\cite{Nagar2005} for more information.}
     \begin{align}
        h_{+} - i h_{\times} & = \frac{1}{r\sqrt{2}}\sum_{l,m} {}_{-2}Y^{lm}\left(\mathcal{Q}^{+}_{lm} -i \int_{-\infty}^t\mathcal{Q}^\times_{lm}(t')dt'\right). \label{eq:Intro:MultipoleExpansion}
    \end{align}   
    Here, $r$ is the distance to the gravitational-wave source and ${}_{-2}Y^{lm}$ are the spin-weighted spherical harmonics with spin $s=-2$ and angular indices of $(l,m)$.
    For most sources, gravitational waves are predominantly emitted in the $l=m=2$ mode. 
    In this case Eq.(\ref{eq:Intro:MultipoleExpansion}) evaluates to
     \begin{align}
        h_{+} + i h_{\times} & = \frac{1}{2 r} \sqrt{\frac{5}{2\pi }} e^{2 i \phi } \cos ^4\left(\frac{\theta }{2}\right)\left(\mathcal{Q}^{+}_{22} -i \int_{-\infty}^t\mathcal{Q}^\times_{22}(t')dt'\right), \label{eq:Intro:MultipoleExpansionlm22}
    \end{align}
    where  $\theta$ is the polar angle and $\phi$ is the azimuthal angle usually defined relative to orbital angular momentum vector. \par

@@

    Assuming a circular orbit of two point masses of mass $M_1$ and $M_2$, the Newtonian quadrupole formula can be used to calculate the evolution of the orbital separation, $a$, due to gravitational wave emission:~\cite[e.g.,][]{Cutler1994}
    \begin{align}
        \frac{da}{dt} & = -\frac{64}{5}\frac{\mu M^2}{a^3}, \label{eq:Intro:OrbitalEvolution}        
    \end{align}
    where  $M \equiv M_1 + M_2$ is the total mass, and $\mu \equiv M_1 M_2 / M$ is the reduced mass.
    The time evolution of gravitational-wave strain frequency can then be calculated as
    \begin{align}
        \frac{df}{dt} & = \frac{96}{5}\pi^{8/3}\mathcal{M}^{5/3}f^{11/3}, \label{eq:Intro:FreqEvolution}        
    \end{align}   
    where $\mathcal{M} \equiv \mu^{3/5} M^{2/5}$ is the chirp mass of the binary system. 
    Using Eqs.(\ref{eq:Intro:OrbitalEvolution}\Hyphdash*\ref{eq:Intro:FreqEvolution}), an expression for the frequency and the gravitational-wave strain magnitude can be calculated~~\cite[e.g.,][]{Cutler1994}: 
      \begin{align}
      f &  \propto \mathcal{M}^{- 5/8}(t_c-t)^{- 3/8},\label{eq:Intro:chirp_f}\\
        |h(t)| & \propto  \frac{\mathcal{M}^{5/4}}{D}(t_c-t)^{-1/4}, \label{eq:Intro:chirp_ht} 
    \end{align}
    where $t$ is the time, $t_c$ is the time until coalescence, and $D$ is the luminosity distance to the source. 
    The combination of Eq.(\ref{eq:Intro:chirp_f})~and~Eq.(\ref{eq:Intro:chirp_ht}) result in the chirp waveform detected in the inspiral of compact object mergers.
    \par
    The first direct gravitational-wave detection from the inspiral of a binary neutron star merger, GW170817, was observed by a detector network consisting of the LIGO detectors at Hanford, Washington and Livingston, Louisiana in the USA, and the Virgo detector in Cascina, Italy~\cite{AdvancedLIGO2015,AdvancedVirgo2015,GW170817Detection}. 
    All three detectors consist of `L' shaped laser interferometers with free-falling test-masses.
    The gravitational-wave strain is found by measuring the optical path difference between the perpendicular beam tubes. \par

    Another binary neutron star (possibly neutron star - black hole) merger, GW190425, was detected in 2019~\cite{GW190425Detection,GWTC2,PopGWTC2}.  However, the luminosity distance of GW190425, $159^{+69}_{-72}\,$Mpc, and the broad sky localisation of the source, $8284\,\mathrm{deg}^2$, precluded any coincident electromagnetic observations~\cite{GW190425Detection}. In comparison, the sky localisation of GW170817 was $28\,\mathrm{deg}^2$ at a distance of $\sim 40\,\Mpc$.
    Therefore, to date, GW170817 is the only gravitational-wave signal that was observed with coincident electromagnetic radiation and will be discussed further in Section~\ref{sec:Intro:NeutronStarMergers}.

@@

% ***************************************************************************************************************************************************************
    \section{Neutron star mergers} \label{sec:Intro:NeutronStarMergers}
    Binary neutron star mergers emit vast amounts of energy in gravitational waves.
    The estimated radiated energy for the GW170817 merger was $\gtrsim 5\times 10^{52}\,$ergs~\cite{GW170817Detection}, equivalent to $\sim 1\%$ of the total rest mass of the two neutron stars.
    For a source at a distance of $D$ from the observer, the amplitude of gravitational waves are  $\propto 1/D$, whereas the amplitude of electromagnetic waves are $\propto 1/D^2$.
    Additionally, electromagnetic waves can be attenuated by intervening media between the source and the detector whereas gravitational waves are transparent to the intervening media.
    Hence gravitational waves are detectable over much larger distances in comparison to electromagnetic radiation with the same source luminosity. \par
    Gravitational-wave strain emitted from the inspiral of binary neutron star mergers, approximated by Eqs.(\ref{eq:Intro:FreqEvolution}\Hyphdash*\ref{eq:Intro:chirp_ht}), can be recast in the frequency domain using the stationary phase approximation~\cite{Cutler1994}
    \begin{align}
        h_{22}(f) & \equiv \frac{\mathcal{A}}{D}\mathcal{M}^{5/6}f^{-7/6} \exp(i\Psi(f)), \label{eq:Intro:StationaryPhaseApproxH22} \\
        \Psi(f) & = 2\pi f t_c - \phi_c - \frac{\pi}{4} +\frac{3}{4}\left(8\pi\mathcal{M}f\right)^{-5/3}, \label{eq:Intro:StationaryPhaseApproxPsi} 
    \end{align}
    where $h_{22}(f)$ is the Fourier transform of the $(l,m)=(2,2)$ gravitational-wave strain, and $\Psi(f)$ is the corresponding phase.
    Here, $f$ is the frequency~(Eq.(\ref{eq:Intro:FreqEvolution})), $\phi_c$ and $t_c$ are the phase and time of coalescence, and $\mathcal{A}$ is a factor that depends on the orientation of the system.\footnote{The orientation factor, $\mathcal{A}$, depends on angles intrinsic and extrinsic to the binary system. Intrinsic angles include the angles defining the relative orientation of the neutron star spins to the orbital plane. Extrinsic angles include the angles defining the relative orientation between the orbital plane and each of the detectors.}
    Equations~\ref{eq:Intro:StationaryPhaseApproxH22}\Hyphdash*\ref{eq:Intro:StationaryPhaseApproxPsi} are valid for $a \gtrsim 6M$ where $a$ is the orbital separation of the binary system and $a\sim 6M$ is an approximation of the innermost stable circular orbit~\cite{Cutler1994}.
    Solving Eqs.(\ref{eq:Intro:OrbitalEvolution}\Hyphdash*\ref{eq:Intro:FreqEvolution}) for $a \sim 6M$ gives $f_{max}\approx c^3 ( 6^{3/2}\pi G M )^{-1}\,\mathrm{Hz}$.
    \par
    % The base post-Newtonian expansion parameter is $\sqrt{M/r}$ but it is more convenient to use $\left(\pi M f\right)^{1/3}$ as the post-Newtonian expansion parameter.
    % These two parameters are equal up to and excluding terms of order $(M/r)^{3/2}$.
    The validity for these equations can be extended to times closer to the merger by adding relativistic corrections calculated by post-Newtonian expansion and effective-one-body dynamics~\cite{Cutler1994,Blanchet1995, Buonanno1999,Droz1999,Damour2012,Yagi2013,Blanchet2014}. % to Eq.~\ref{eq:Intro:StationaryPhaseApproxPsi}
    These corrections introduce dependence on the masses, spins, and tidal deformabilities of the individual neutron stars.
    This allows the measurement of these system properties in addition to the chirp mass, given a gravitational-wave inspiral detection.
    However, the accuracy of the individual neutron star properties is significantly less than the accuracy of the chirp mass because the relativistic corrections required to measure these properties are fairly weak. 
    This can be seen by looking at the parameters measured from GW170817.

@@

    \section{Numerical relativity simulations}

    Numerical-relativity simulations are required for determining the dynamics of the late inspiral of the binary neutron star and the post-merger remnant.
    To perform a numerical-relativity simulation, the equation of state must be defined, a set of initial conditions must be evaluated in general relativity, and then Einstein's field equations must be evolved.
    Each of these three requirements make these simulations difficult.
    The equation of state for a `normal' neutron star is still unknown, and the equation of state for a hot post-merger remnant is even more uncertain, after all, this is the hottest and densest bulk matter in the Universe.\par

    The initial conditions for binary neutron star mergers in general relativity are complex and a number of methods have been developed for this task.
    Codes that have been used for this include \texttt{LORENE}~\cite{Gourgoulhon2001} and \texttt{COCAL}~\cite{Tsokaros2015}.
    The \texttt{LORENE} formulation assumes a conformally-flat approximation for  neutron star binaries and uses spectral decomposition to solve for the initial conditions~\cite{Gourgoulhon2001}.
    This formulation constructs quasiequilibrium configurations which are approximations of the inspiral dynamics using exactly circular orbits.
    Quasiequilibrium configurations are also generated by \texttt{COCAL}.
    This is performed by excising regions around the neutron stars and solving Green's integral formula using multipole expansion and finite-difference methods~\cite{Uryu2012,Tsokaros2015}.
    Once the initial conditions have been approximated then the space-time can be evolved.

    Space-time evolution methods include the following formulations: Arnowitt-Deser-Misner (ADM)~\cite{Arnowitt2008}\footnote{The original book reference for ADM formalism is out of print and the authors have listed the cited paper as an `intentially unretouched' alternative.},  Baumgarte-Shapiro-Shibata-Nakamura-Oohara-Kojima (BSSNOK)~\cite{Nakamura1987,Shibata1995,Baumgarte1999,Alcubierre2000} and Z4c~\cite{Bona2003,Bona2004,Gundlach2005,Bernuzzi2010,Hilditch2013}. 
    The ADM formalism is a Hamiltonian method that splits the Einstein's field equations into separate time and space components.
    This is achieved by applying a gauge that consists of a shift vector and lapse function~\cite{Arnowitt2008}.
    Unfortunately, ADM numerical simulations were found to be unstable to accumulated errors over the simulation time~\cite{Brandt2000,Kidder2001,Alcubierre2001a} and a number of different formulations were developed to improve this. \par

@@

    \section{Post-merger remnants} \label{sec:Intro:PMR}
